Convergence rate of MAP estimates \\ for the exponential family

### 2021-06-18 main.tex
Move 3 point lemma to main text
Declutter gaussian example and bias-variance decomposition sections by moving to appendix

\newmdenv[% box with grey background
	style=blurry-shadow,
    backgroundcolor=gray!15,
]{example}
\newmdenv[
	style=blurry-shadow,
	backgroundcolor=myblue,
]{important}

@@

%% Faster processing
\renewcommand\appendix{\end{document}}

@@

\paragraph{Conjugate Prior}
The conjugate prior for $p(X|\natp)$ is
\begin{align}
    \pi(\natp) &\propto \exp( - n_0 \bregman(\natp ; \natp_0) )
\end{align}
where $n_0$ and $\natp_0$ are (hyper)parameters of the prior, and $\bregman(\natp ; \natp_0)$ is the Bregman divergence induced by $\logpart$ between $\natp$ and $\natp_0$
\begin{align}
    \bregman (\natp ; \natp_0)
    & = \logpart(\natp) - \logpart(\natp_0) 
    - \langle \nabla \logpart(\natp_0)  , \natp - \natp_0 \rangle
\end{align}
with $\nabla \logpart(\natp_0) = \expect[\natp_0]{T(X)} =: \meanp_0$ the mean parameter associated to $\natp_0$. 
Intuitively, $n_0$ is a number of fictive points observed from a distribution with parameter $\natp_0$.
We can re-write this prior as 
\begin{align}
    \pi(\natp) \propto 
    \exp( -n_0 \logpart (\natp) 
    + \langle n_0 \mu_0, \natp \rangle ) \; ,
\end{align}
which is the formula for the exponential family with sufficient statistics $(\natp ,\logpart(\natp))$ and with natural parameter $(n_0 \mu_0, -n_0)$.
The posterior given $\cD=(X_1,\dots, X_n)$ is then part of the same family, with natural parameters $(n_0 \mu_0 + \sum_i T(X_i) , -(n_0 + n))$.

@@

\section{Problem}
For a well-specified model, the suboptimality on the population log-likelihood is exactly the KL between our current model and the true distribution
\begin{align}
    \expect[X\sim p(.|\natp^*)]{-\log p(X|\natp) + \log p(X|\natp^*) }
	= \KL( p(.|\natp^*) ; p(.|\natp)) \; .
\end{align}
For the exponential family, the KL is also the Bregman divergence induced by the log-partition function (with switched arguments)
\begin{align}
	\KL( p(.|\natp^*) ; p(.|\natp)) 
	= \bregman (\natp ; \natp^*)  \; .
\end{align}
There is a general relationship between Bregman divergences and convex conjugates (notice the argument switching)
\begin{align}
	\bregman (\natp ; \natp^*)
    = \logpart(\natp) - \langle \natp , \mu^* \rangle + \conj(\mu^*)
    = \bregmanconj ( \meanp^* ; \meanp)
\end{align}
so in the end the suboptimality is a divergence, which can either be seen as a KL between distributions, as a divergence between natural parameters, or as a divergence between mean parameters
\begin{align}
\boxed{
	\KL( p(.|\natp^*) ; p(.|\natp))
    = \bregman (\natp ; \natp^*)
    = \bregmanconj ( \meanp^* ; \meanp) \; .
}
\end{align}
\begin{important}
The question is: how does this quantity behave when $\natp$ is the maximum-likelihood or the MAP estimate ? Can we get bounds on the following quantities
\begin{align}
	\label{eq:bregmanMLE}
	\expect[X_i\sim \natp^*]{\bregmanconj \left (\E [T(X)] ;  \inv{n}  \smallsum_i T(X_i) \right )} \leq \ ? \; , \\
	\label{eq:bregmanMAP}
	\expect[X_i\sim \natp^*]{\bregmanconj \left (\E [T(X)] ; \frac{n_0 \mu_0 + \smallsum_i T(X_i)}{n_0+n} \right )} \leq \ ? \; ,
\end{align}
where the outer expectation is on the dataset $X_1, \dots, X_n$?
\end{important}

@@

\begin{example}
	For $\cN(0,\sigma^2)$, both the entropy and the log-partition are roughly negative logarithm $z\mapsto - \log(z)$. Which yields the same shape of Bregman divergence, as visible below (all three lines are equal)
\begin{align}
	\KL( \sigma_*^2 ; \sigma_n^2 ) 
	&= \half \left ( \frac{\sigma_*^2}{ \sigma_n^2} - 1 - \log \frac{\sigma_*^2}{ \sigma_n^2} \right) \\
	\bregmanconj( \mu_*; \mu_n) 
	&= \half \left ( \frac{\mu_*}{ \mu_n} - 1 - \log  \frac{\mu_*}{ \mu_n} \right) \\
	\bregman( \natp_n; \natp_* ) 
	&=  \half \left ( \frac{ \natp_n}{\natp_*} - 1 - \log  \frac{ \natp_n}{\natp_*} \right) \; .
\end{align}
In other words, this divergence measures the discrepancy between the ratio $\frac{ \natp_n}{\natp_*} =  \frac{\mu_*}{ \mu_n}  $ and $1$ via the function $\phi$
\begin{align}
	\phi(z) := \half (z - 1 - \log(z)) \\
	\bregman( \natp_n; \natp_* )   = \phi(\frac{ \natp_n}{\natp_*}) =  \phi(\frac{\mu_*}{ \mu_n})
\end{align}
as illustrated in Figure~\ref{fig:phi}. We can get a non-transcendental upper bound thanks to the inequality
\begin{align}
	1 - \inv{z} &\leq \log(z) 
	\label{eq:log_bound}\\
	\implies \phi(z) &\leq \half (z + \inv{z}) - 1 = \frac{(z-1)^2}{2 z}
\end{align}
as illustrated in Figure~\ref{fig:phi}.
\end{example}

\begin{figure}[ht]
	\centering
	\includegraphics[width=.4\textwidth]{figs/bregmandef.pdf}
	\includegraphics[width=.4\textwidth]{figs/phi.pdf}
	\caption{$\phi(z)$ is the Bregman divergence induced by $-\log(z)$. It is a barrier near $0$, so it is poorly approximated by quadratics.}
	\label{fig:phi}
\end{figure}

@@

\begin{example}
	\begin{theorem}[MLE tight upper bound]
	The MLE of $\cN(0,\mu_*)$ is $\hat \mu_n^\text{MLE} = \inv{n} \sum_i X_i^2 $.
	Its expected suboptimality is infinite when $n\leq 2$, and otherwise upper-bounded as
	\begin{align}
		 \expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MLE}) }
			\leq \inv{2n} +\frac{2}{n(n-2)} \; .
			\label{eq:MLE_rate}
	\end{align}
\end{theorem}
Note that this upper bound is asymptotically tight.
 There is also a closed form for the multivariate generalization, thanks to the \href{https://en.wikipedia.org/wiki/Inverse-Wishart_distribution}{inverse Wishart distribution} and the \href{https://en.wikipedia.org/wiki/Wishart_distribution#Log-expectation}{expectation of the log-determinant of a Wishart}. The expected value is infinite whenever $n \leq d+1$ where $d$ is the dimension. We do not perform these calculations for now as we want to focus on the simplest possible examples.

As for the MAP, we did not manage to get an asymptotically tight upper bound. Nevertheless, using inequality~\eqref{eq:log_bound}, we did get an interesting upper bound. 
To start, let us recall that the MAP of $\cN(0,\mu_*)$ is $\hat \mu_n = \frac{n_0 \mu_0 + \sum_i X_i^2}{n_0 + n}$.
Its expectation is simply $\mu_n= \frac{n_0 \mu_0 + n \mu^*}{n_0 + n}$.
We now present a lemma on its expected inverse.
\begin{lemma}[Expected MAP natural parameter]
	For any $n\geq 2$, the expectation of the natural parameter of the MAP of $\cN(0,\mu_*)$ is bounded as
	\begin{align}
		\frac{\mu^*}{\mu_n}
		\leq \expect{\frac{\mu^*}{\MAPm}} 
		= \expect{\frac{\MAPt}{\natp^*}} 
		\leq \frac{1}{\frac{\mu_n}{\mu^*} - \frac{2}{n+n_0}} \; .
	\end{align}
\end{lemma}
Not that the lower bound is a simple consequence of the convexity of the inverse function. Using this lemma, along withthe los upper bound on the logarithm~\eqref{eq:log_bound}, we get a convergence rate for the MAP. We relate this rate convergence rate with the rate of MLE in Figure~\ref{fig:MAP_rate}.
\begin{theorem}[MAP Upper Bound]
 Introducing $a=n_0 \frac{\mu_0}{\mu^*}$, the expected suboptimality of the MAP of $\cN(0,\mu^*)$ with prior hyper-parameters $(n_0,\mu_0)$ is 
\begin{equation}
	\expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MAP})}
	\leq \begin{cases}
		\bregmanconj( \mu_*; \mu_0) \ \text{ if } \ n=0, \\
		\bregmanconj( \mu_*; \mu_0) + \inv{a} \ \text{if}\ n=1,\\
		\frac{1}{a+n-2} + \frac{\half(1 - \frac{\mu_0}{\mu^*})^2}{(\frac{\mu_0}{\mu^*}+\frac{n-2}{n_0})(1 + \frac{n}{n_0} )} \ \text{if}\  n\geq 2 \ \text{and}\ a=n_0 \frac{\mu_0}{\mu^*} >0
	\end{cases}
	\label{eq:MAP_rate}
\end{equation}
\end{theorem}
Using the same bound on the log~\eqref{eq:log_bound}, we can also get a loose bound on the MLE, which is useful for comparison purposes.
\begin{equation}
	\expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MLE}) }
	\leq \inv{n-2} \; .
	\label{eq:MLE_loose_rate}
\end{equation}

\end{example}

@@

\section{Proofs for Gaussian Variance}
\subsection{MLE Tight Upper Bound}
	\begin{theorem}[MLE tight upper bound]
	The MLE of $\cN(0,\mu_*)$ is $\mu_n = \inv{n} \sum_i X_i^2 $.
	Its expected suboptimality is infinite when $n\leq 2$ and otherwise
	\begin{align}
		 \expect{\bregmanconj( \mu_*; \mu_n) }
			\leq \inv{2n} +\frac{2}{n(n-2)} \; .
	\end{align}
	\end{theorem}

\subsection{MLE Loose Upper Bound}
If we apply \eqref{eq:log_bound} to the MLE, we get the upper bound
\begin{align}
	\expect{\bregmanconj(\mu^*; \MAPm^\text{MLE})}
	&\leq  \half \left( \expect{\frac{n}{\chi^2(n)}} + \expect{\frac{\chi^2(n)}{n}} - 2 \right ) \\
	&= \half \left( \frac{n}{n-2} + 1 - 2 \right)  
	= \frac{1}{n-2}
\end{align}
which is exactly  an additive term $1/2n$ larger than \eqref{eq:MLE_rate}. This highlights the factor $2$ difference between the log-bound and an exact computation of the expectation of the log term, and a tight upper bound on the digamma function.

\subsection{Expected MAP Natural Parameter}
\begin{lemma}[Expected MAP natural parameter]
	The expectation of the natural parameter of the MAP of $\cN(0,\mu_*)$ is bounded as
	\begin{align}
		\frac{\mu^*}{\mu_n}
		\leq \expect{\frac{\mu^*}{\MAPm}} 
		= \expect{\frac{\MAPt}{\natp^*}} 
		\leq \frac{1}{\frac{\mu_n}{\mu^*} - \frac{2}{n+n_0}} \; .
	\end{align}
\end{lemma}

\begin{proof}
	To start, let us plug in the definition of $\MAPm$
	\begin{align}
		\expect{\frac{\mu^*}{\MAPm}} 
		= \expect{\frac{\MAPt}{\natp^*}} 
		&= \expect{\frac{(n_0+n) \mu^*}{n_0 \mu_0 + \sum_i X_i^2}}  \\
		&= (n_0 + n) \expect{\inv{n_0 \frac{\mu_0}{\mu^*}+ \sum_i \frac{X_i^2}{ \mu^*} }} \\
		&= (n_0 +n) \expect{\inv{a + \chi^2(n)}}
	\end{align}
	where $\chi^2(n) = \sum_i \frac{X_i^2}{ \mu^*}$ is a chi-square random variable of degree $n$ and 
\begin{important}
	\begin{align}
			a=n_0 \frac{\mu_0}{\mu^*} \; .
	\end{align}
\end{important}
	We reformulate this expectation with the trick
	\begin{align}
		&\int_1^\infty e^{-\frac{zt}{2}} dt = \frac{2}{z} e^{-\half[z]} \\
		\implies &\inv{z} = \half e^{\half[z]} \int_1^\infty e^{-\frac{zt}{2}} dt \\
		\implies &\inv{a+x} = \half e^{\frac{a+x}{2}} \int_1^\infty e^{- \frac{(a+x)t}{2}}dt \; ,
	\end{align} 
	to get two integrals instead of one
	\begin{align}
		2^{\half[n]}\Gamma(\half[n]) \expect{\inv{a + \chi^2(n)}}  
		&= \int_0^\infty \frac{x^{\frac{n}{2}-1}e^{-\half[x]} }{a+x} dx \\
		&= \int_0^\infty dx\  x^{\frac{n}{2}-1}e^{-\half[x]} \half e^{\half[a+x]} \int_1^\infty  dt\  e^{-\half[(a+x)t]}  \\
		&= \half e^{\half[a]} \int_1^\infty dt \ e^{-\half[a t]}   \int_0^\infty dx \  x^{\frac{n}{2}-1} e^{-\half[x t]}  \; ,
	\end{align}
	where we used Fubini to switch integrals, without further justification.
	Now with the change of variable $x = 2\frac{y}{t}$, eg $dx = 2\frac{dy}{t}$, the inner integral becomes
	\begin{align}
		\expect{\inv{a + \chi^2(n)}}  
		&= \frac{e^{\half[a]} }{2^{\half[n]+1}\Gamma(\half[n]) }\int_1^\infty dt \ e^{-\half[a t]}   \int_0^\infty 2\frac{dy}{t} \  (2\frac{y}{t})^{\frac{n}{2}-1} e^{-y} \\
		&= \frac{e^{\half[a]} }{2\Gamma(\half[n]) } 
		\underbrace{\int_1^\infty  \frac{e^{-\half[a t]}}{t^{\frac{n}{2}}}  dt}_{E_{\half[n]}(\half[a])  } \ 
		\underbrace{\int_0^\infty  y^{\frac{n}{2}-1} e^{-y} dy}_{\Gamma(\half[n])}
	\end{align}
	so we finally get  the formula valid for all $n\geq 1$
	\begin{align}
		\boxed{\expect{\inv{a + \chi^2(n)}}  = \half e^{\frac{a}{2}} E_{\half[n]}(\half[a])}
		\label{eq:hard_expectation}
	\end{align}
	where the \href{https://dlmf.nist.gov/8.19}{generalized exponential integral function} is defined as 
	\begin{align}
		E_k(z) = \int_1^\infty \frac{e^{-z t} }{t^k} dt \; .
	\end{align}
	Now our goal is to bound this function with simpler functions.
	Fortunately, mathematicians have been working on these integrals for decades.
	When $n=2$, eg $k=1$, we have the simple exponential integral function which \href{https://dlmf.nist.gov/6.8\#E1}{verifies the bound}
	\begin{align}
		\half \log(1 + \frac{2}{x}) 
		\leq e^x E_1(x) 
		\leq \log(1+\inv{x}) \leq \inv{x} \; .
	\end{align}
	For $n \geq 2$, eg $k\geq 1$, we have \href{https://dlmf.nist.gov/8.19\#E21}{the general bound}
	\begin{align}
		\inv{x+k}
		\leq e^x E_k(x) 
		\leq \inv{x + k - 1}
	\end{align}
	so that 
	\begin{align}
		\inv{a+n}
		&\leq \expect{\inv{a + \chi^2(n)}}
		\leq \inv{a + n - 2} \\
		\iff \frac{n +n_0}{a + n}	
		&\leq  \expect{\frac{\mu^*}{\MAPm}} 
		\leq \frac{n +n_0}{a + n - 2}	\\
		\iff \frac{\mu^*}{\mu_n}
		&\leq \expect{\frac{\mu^*}{\MAPm}} 
		\leq \frac{1}{\frac{\mu_n}{\mu^*} - \frac{2}{n + n_0}}  
		\label{eq:theta_expectation_bound}
	\end{align}
	where we used $\frac{a + n}{n +n_0} = \frac{\mu_n}{\mu^*}$.
\end{proof}

\subsection{MAP Upper Bound}
\begin{theorem}[MAP Upper Bound]
 The expected suboptimality of the MAP of $\cN(0,\mu^*)$ with prior hyper-parameters $(n_0,\mu_0)$ is 
\begin{equation}
	\expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MAP})}
	\leq \begin{cases}
		\bregmanconj( \mu_*; \mu_0) \ \text{ if } \ n=0, \\
		\bregmanconj( \mu_*; \mu_0) +\inv{a} \ \text{if}\ n=1,\\
		\frac{1}{a+n-2} + \frac{\half(1 - \frac{\mu_0}{\mu^*})^2}{(\frac{\mu_0}{\mu^*}+\frac{n-2}{n_0})(1 + \frac{n}{n_0} )} if n\geq 2 \ \text{and}\ a=n_0 \frac{\mu_0}{\mu^*} >0
	\end{cases}
\end{equation}
\end{theorem}

\begin{proof}
		Wolfram finds no closed form for $\expect{\log\frac{\MAPm}{\mu^*}}$ so we focus solely on the simpler upper bound given Now using the bound \eqref{eq:log_bound} on the logarithm, we get
\begin{align}
	2\expect{\bregmanconj(\mu^*; \MAPm)} 
	&\leq \expect{\frac{\mu^*}{\MAPm}}  + \expect{\frac{\MAPm}{\mu^*}} - 1 \\
	&\leq \frac{\mu^*}{\mu_n - \frac{2 \mu^*}{n + n_0}}   + \frac{\mu_n}{\mu^*} - 1 \\
	&= \frac{n_0 + n}{a +n - 2} + \frac{a+n}{n_0 + n} - 2 \\
	&= \frac{n_0 - a + 2}{a +n - 2} + \frac{a - n_0}{n_0 + n} \\
	&= \frac{2}{a+n-2} + \frac{(n_0 - a)^2}{(a+n-2)(n_0 + n)} \\
	\implies \expect{\bregmanconj(\mu^*; \MAPm)} 
	&\leq \frac{1}{a+n-2} + \frac{(n_0 - a)^2}{2(a+n-2)(n_0 + n)} \\
	&\leq \frac{1}{a+n-2} + \frac{\half(1 - \frac{\mu_0}{\mu^*})^2}{(\frac{\mu_0}{\mu^*}+\frac{n-2}{n_0})(1 + \frac{n}{n_0} )} 
\end{align}
where recall that $a=n_0\frac{\mu_0}{\mu^*}$. 
This inequality highlights a clear variance-bias decomposition, to compare with the MLE. Unfortunately it is not tight asymptotically as the inequality \eqref{eq:log_bound} is not quadratically tight around $1$. We are basically losing a factor 2 compared to $1/2n$. by~\eqref{eq:log_bound}.
\end{proof}

\section{Comparison with  Euclidean Distance}
Often we can relate this Bregman suboptimality with the $\ell^2$ distance in mean parameter space.

\paragraph{If $\conj$ is $L$-Lipschitz} (e.g. $\logpart$ is defined within the $\ell^2$-ball of radius $L$), then
\begin{align}
    \bregmanconj(\mu^* ; \mu) 
    \leq L \norm{\mu^* - \mu} + \norm{\natp} \norm{\mu^* - \mu}
    \leq 2L \norm{\mu^* - \mu}
\end{align}
so $\bregmanconj$ is $2L$-Lipschitz.
Since the empirical average converges in expectation to the population mean at a rate of $1/\sqrt{n}$ in $\ell^2$ norm, we know that this bound applies to the log-likelihood.

\paragraph{If $\conj$ is $L$-smooth} (e.g. $\logpart$ is $L^{-1}$-strongly convex), then
\begin{align}
    \bregmanconj(\mu^* ; \mu) 
    \leq \frac{L}{2} \norm{\mu^* - \mu}^2
\end{align}
so $\bregmanconj$ is upper bounded by a quadratic. In expectation, it should converge at a rate $1/n$.

@@

\subsection{Pure Bias Term}
The pure bias term measures the divergence between the optimum and the expected MAP estimate $\bregmanconj(\mu^*; \mu_n)$.

\paragraph{Aymptote.}
Using the quadratic approximation of the Bregman divergence near $\mu^*$, we get a $O(n^{-2})$ rate for the pure bias term
\begin{align}
	\bregmanconj(\mu^*; \mu_n) 
	= \frac{\norm{\mu^* -  \mu_0}^2_{\mSigma^{-1}}}{2(1 + \frac{n}{n_0})^{-2} } + O(n^{-3}) \; .
\end{align}

\subsection{Pure Variance Term}
The pure variance term $\Cov(\hat \natp_n ; \hat \mu_n )$   can be seen as a non-linear variance of $\hat \mu_n$.
Indeed, it is positive thanks to the monotonicity of $\nabla \conj$
\begin{align}
&\langle \hat \natp_n - \natp_n ; \hat \mu_n - \mu_n \rangle 
	= \langle \nabla \conj (\hat \mu_n) - \nabla \conj(\mu_n) ;  \hat \mu_n - \mu_n \rangle 
	\geq 0
\end{align}
Taking the expectation, (which allows removing the constant $\theta_n$) 
\begin{align}
	\expect{\langle \hat \natp_n - \natp_n ; \hat \mu_n - \mu_n \rangle} 
	= \Cov(\hat \natp_n ; \hat \mu_n ) \geq 0
\end{align}
Consequently, we will now write about this covariance as a variance, with the monotone operator $\nabla \conj$ (or equivalently $\nabla \logpart$) plugged in the middle
\begin{align}
	\Var_{\nabla \conj}(\hat \mu_n) := \Cov(\nabla\conj(\hat \mu_n) ; \hat \mu_n ) \geq 0 \; .
\end{align}

>>>>

\newmdenv[% box with grey background
	style=blurry-shadow,
    backgroundcolor=gray!15,
]{example}
\newmdenv[
	style=blurry-shadow,
	backgroundcolor=Red!30,
]{important}

@@

%% Faster processing
%\renewcommand\appendix{\end{document}}

@@

\paragraph{Conjugate Prior}
\TODO{Under-parametrized conjugate prior ?}
The conjugate prior for $p(X|\natp)$ is
\begin{align}
    \pi(\natp) &\propto \exp( - n_0 \bregman(\natp ; \natp_0) )
\end{align}
where $n_0$ and $\natp_0$ are (hyper)parameters of the prior, and $\bregman(\natp ; \natp_0)$ is the Bregman divergence induced by $\logpart$ between $\natp$ and $\natp_0$
\begin{align}
    \bregman (\natp ; \natp_0)
    & = \logpart(\natp) - \logpart(\natp_0) 
    - \langle \nabla \logpart(\natp_0)  , \natp - \natp_0 \rangle
\end{align}
with $\nabla \logpart(\natp_0) = \expect[\natp_0]{T(X)} =: \meanp_0$ the mean parameter associated to $\natp_0$. 
Intuitively, $n_0$ is a number of fictive points observed from a distribution with parameter $\natp_0$.
We can re-write this prior as 
\begin{align}
    \pi(\natp) \propto 
    \exp( -n_0 \logpart (\natp) 
    + \langle n_0 \mu_0, \natp \rangle ) \; ,
\end{align}
which is the formula for the exponential family with sufficient statistics $(\natp ,\logpart(\natp))$ and with natural parameter $(n_0 \mu_0, -n_0)$.
The posterior given $\cD=(X_1,\dots, X_n)$ is then part of the same family, with natural parameters $(n_0 \mu_0 + \sum_i T(X_i) , -(n_0 + n))$.

@@

\section{Open Problem}
For a well-specified model, the suboptimality on the population log-likelihood is exactly the KL between our current model and the true distribution
\begin{align}
    \expect[X\sim p(.|\natp^*)]{-\log p(X|\natp) + \log p(X|\natp^*) }
	= \KL( p(.|\natp^*) ; p(.|\natp)) \; .
\end{align}
For the exponential family, the KL is also the Bregman divergence induced by the log-partition function (with switched arguments)
\begin{align}
	\KL( p(.|\natp^*) ; p(.|\natp)) 
	= \bregman (\natp ; \natp^*)  \; .
\end{align}
There is a general relationship between Bregman divergences and convex conjugates (notice the argument switching)
\begin{align}
	\bregman (\natp ; \natp^*)
    = \logpart(\natp) - \langle \natp , \mu^* \rangle + \conj(\mu^*)
    = \bregmanconj ( \meanp^* ; \meanp)
\end{align}
so in the end the suboptimality is a divergence, which can either be seen as a KL between distributions, as a divergence between natural parameters, or as a divergence between mean parameters
\begin{align}
\boxed{
	\KL( p(.|\natp^*) ; p(.|\natp))
    = \bregman (\natp ; \natp^*)
    = \bregmanconj ( \meanp^* ; \meanp) \; .
}
\end{align}
\begin{important}
The question is: how does this quantity behave when $\natp$ is the maximum-likelihood or the MAP estimate ? Can we get bounds on the following quantities
\begin{align}
	\label{eq:bregmanMLE}
	\expect[X_i\sim \natp^*]{\bregmanconj \left (\E [T(X)] ;  \inv{n}  \smallsum_i T(X_i) \right )} \leq \ ? \; , \\
	\label{eq:bregmanMAP}
	\expect[X_i\sim \natp^*]{\bregmanconj \left (\E [T(X)] ; \frac{n_0 \mu_0 + \smallsum_i T(X_i)}{n_0+n} \right )} \leq \ ? \; ,
\end{align}
where the outer expectation is on the dataset $X_1, \dots, X_n$?
\end{important}

@@

\paragraph{Remark 2.}
The expectation of the MLE may be infinite, for instance with $\cN(0,\sigma^2)$ and $n\leq 2$. Instead of taking the expectation,  we might want to bound this quantity in high probability, without resorting to Markov inequality, but that is a difficult endeavor.

Below we review all the attempts we made on this problem.

\section{Gaussian Variance Example}

\begin{example}
	For $\cN(0,\sigma^2)$, both the entropy and the log-partition are roughly negative logarithm $z\mapsto - \log(z)$. Which yields the same shape of Bregman divergence, as visible below (all three lines are equal)
\begin{align}
	\KL( \sigma_*^2 ; \sigma_n^2 ) 
	&= \half \left ( \frac{\sigma_*^2}{ \sigma_n^2} - 1 - \log \frac{\sigma_*^2}{ \sigma_n^2} \right) \\
	\bregmanconj( \mu_*; \mu_n) 
	&= \half \left ( \frac{\mu_*}{ \mu_n} - 1 - \log  \frac{\mu_*}{ \mu_n} \right) \\
	\bregman( \natp_n; \natp_* ) 
	&=  \half \left ( \frac{ \natp_n}{\natp_*} - 1 - \log  \frac{ \natp_n}{\natp_*} \right) \; .
\end{align}
\end{example}
In other words, this divergence measures the discrepancy between the ratio $\frac{ \natp_n}{\natp_*} =  \frac{\mu_*}{ \mu_n}  $ and $1$ via the function $\phi$
\begin{align}
	\phi(z) := \half (z - 1 - \log(z)) \\
	\bregman( \natp_n; \natp_* )   = \phi(\frac{ \natp_n}{\natp_*}) =  \phi(\frac{\mu_*}{ \mu_n})
\end{align}
as illustrated in Figure~\ref{fig:phi}. We can get a non-transcendental upper bound thanks to the inequality
\begin{align}
	1 - \inv{z} \leq \log(z) 
	\label{eq:log_bound} 
	\implies \phi(z) \leq \half (z + \inv{z}) - 1 = \frac{(z-1)^2}{2 z} \; .
\end{align}

\begin{figure}[ht]
	\centering
	\includegraphics[width=.4\textwidth]{figs/bregmandef.pdf}
	\includegraphics[width=.4\textwidth]{figs/phi.pdf}
	\caption{$\phi(z)$ is the Bregman divergence induced by $-\log(z)$. It is a barrier near $0$. As a result, it is poorly approximated by quadratics.}
	\label{fig:phi}
\end{figure}

@@

\begin{example}
	\begin{theorem}[MLE Tight Bound]
	The MLE of $\cN(0,\mu_*)$ is $\hat \mu_n^\text{MLE} = \inv{n} \sum_i X_i^2 $.
	Its expected suboptimality is infinite when $n\leq 2$, and otherwise upper-bounded as
	\begin{align}
		 \expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MLE}) }
			\leq \inv{2n} +\frac{2}{n(n-2)} \; .
			\label{eq:MLE_rate}
	\end{align}
\end{theorem}
\end{example}
This upper bound is asymptotically tight.
We illustrate its behavior against empirical data in Figure~\ref{fig:curves}.
 There is also a closed form for the multivariate generalization, thanks to the \href{https://en.wikipedia.org/wiki/Inverse-Wishart_distribution}{inverse Wishart distribution} and the \href{https://en.wikipedia.org/wiki/Wishart_distribution#Log-expectation}{expectation of the log-determinant of a Wishart}. The expected value is infinite whenever $n \leq d+1$ where $d$ is the dimension. We do not perform these calculations for now as we want to focus on the simplest possible examples.

As for the MAP, we did not manage to get an asymptotically tight upper bound. Nevertheless, using inequality~\eqref{eq:log_bound}, we did get an interesting upper bound. 
To start, let us recall that the MAP of $\cN(0,\mu_*)$ is $\hat \mu_n = \frac{n_0 \mu_0 + \sum_i X_i^2}{n_0 + n}$.
Its expectation is simply $\mu_n= \frac{n_0 \mu_0 + n \mu^*}{n_0 + n}$.
We now present a lemma on its expected inverse.
\begin{example}
\begin{lemma}[Expected MAP Natural Parameter]
	For any $n\geq 2$, the expectation of the natural parameter of the MAP of $\cN(0,\mu_*)$ is bounded as
	\begin{align}
		\frac{\mu^*}{\mu_n}
		\leq \expect{\frac{\mu^*}{\MAPm}} 
		= \expect{\frac{\MAPt}{\natp^*}} 
		\leq \frac{1}{\frac{\mu_n}{\mu^*} - \frac{2}{n+n_0}} \; .
	\label{eq:theta_expectation_bound}
	\end{align}
\end{lemma}
\end{example}
Note that the lower bound is a simple consequence of the convexity of the inverse function. 
Using this lemma, along with the log upper bound on the logarithm~\eqref{eq:log_bound}, we get a convergence rate for the MAP. 
\begin{example}
\begin{theorem}[MAP Bound]
 Introducing $a=n_0 \frac{\mu_0}{\mu^*}$, the expected suboptimality of the MAP of $\cN(0,\mu^*)$ with prior hyper-parameters $(n_0,\mu_0)$ is 
\begin{equation}
	\expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MAP})}
	\leq \begin{cases}
		\bregmanconj( \mu_*; \mu_0) \ \text{ if } \ n=0, \\
		\bregmanconj( \mu_*; \mu_0) + \inv{a} \ \text{if}\ n=1,\\
		\frac{1}{a+n-2} + \frac{\half(1 - \frac{\mu_0}{\mu^*})^2}{(\frac{\mu_0}{\mu^*}+\frac{n-2}{n_0})(1 + \frac{n}{n_0} )} \ \text{if}\  n\geq 2 \ \text{and}\ a=n_0 \frac{\mu_0}{\mu^*} >0
	\end{cases}
	\label{eq:MAP_rate}
\end{equation}
\end{theorem}
\end{example}
Using the same bound on the log~\eqref{eq:log_bound}, we  get a loose bound on the MLE, which is useful for comparison purposes.
We relate these loose convergence rates of MAP and MLE in Figure~\ref{fig:MAP_rate}.
\begin{example}
\begin{corollary}[MLE Loose Bound]
\begin{equation}
	\expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MLE}) }
	\leq \inv{n-2} \; .
	\label{eq:MLE_loose_rate}
\end{equation}
\end{corollary}
\end{example}

@@

\section{Euclidean Behaviour}
Often we can relate this Bregman suboptimality with the $\ell^2$ distance in mean parameter space. 
\TODO{Add base measure to add here a discussion of the euclidean case -- the mean of a gaussian}
\paragraph{If $\conj$ is $L$-Lipschitz} (e.g. $\logpart$ is defined within the $\ell^2$-ball of radius $L$), then
\begin{align}
    \bregmanconj(\mu^* ; \mu) 
    \leq L \norm{\mu^* - \mu} + \norm{\natp} \norm{\mu^* - \mu}
    \leq 2L \norm{\mu^* - \mu}
\end{align}
so $\bregmanconj$ is $2L$-Lipschitz.
Since the empirical average converges in expectation to the population mean at a rate of $1/\sqrt{n}$ in $\ell^2$ norm, we know that this bound applies to the log-likelihood.
\paragraph{If $\conj$ is $L$-smooth} (e.g. $\logpart$ is $L^{-1}$-strongly convex), then
\begin{align}
    \bregmanconj(\mu^* ; \mu) 
    \leq \frac{L}{2} \norm{\mu^* - \mu}^2
\end{align}
so $\bregmanconj$ is upper bounded by a quadratic. In expectation, it should converge at a rate $1/n$.

@@

\subsection{About Variance and Covariance}

\begin{lemma}
Taking  $\MAPt$ and $\MAPm$ as the primal and dual parameters of the MAP estimate, and $\tilde \mu_n$ the primal expectation and $\mu_n$ the dual expectation, the following identities hold
	\begin{align}
	\Cov(\hat \natp_n ; \MAPm )  
	&=\expect{\bregmanconj(\MAPm ; \mu_n)
	+ \bregmanconj(\mu_n ; \MAPm)} \\
	&=\expect{\bregmanconj(\MAPm ; \tilde \mu_n)
	+ \bregmanconj(\tilde \mu_n ; \MAPm)} \; .
\end{align}
\end{lemma}
\begin{proof}
By the three point identity (or plain derivation), we have 
\begin{align}
	\bregmanconj(\MAPm ; \mu_n)
	+ \bregmanconj(\mu_n ; \MAPm)
	= \langle \nabla \conj (\hat \mu_n) - \nabla \conj(\mu_n) ;  \hat \mu_n - \mu_n \rangle 
	= \langle \MAPt - \theta_n;  \MAPm- \mu_n \rangle 
\end{align}
Taking the expectation,  the constant $\theta_n$ disappears since $\expect{\hat \mu_n - \mu_n}=0$.
\begin{align}
	\expect{\langle \MAPt - \theta_n;  \MAPm- \mu_n \rangle } 
	 = \expect{\langle \MAPt; \MAPm \rangle} - \langle \expect{\MAPt} ; \MAPm \rangle
	= \Cov(\hat \natp_n ; \hat \mu_n ) 
\end{align}
which is the first identity. We get the second identity by symmetry between the primal and the dual.
\end{proof}

Note that, as a corollary, the covariance between primal and dual MAP
$\Cov(\hat \natp_n ; \hat \mu_n )$ is positive, and it can be seen as a non-linear variance of $\hat \mu_n$ -- e.g. we can define a variance operator with a non-linear monotone operator $\nabla\conj$ plugged in the middle $\Var_{\nabla \conj}(\hat \mu_n) := \Cov(\nabla\conj(\hat \mu_n) ; \hat \mu_n ) \geq 0$.

\paragraph{Expansion}
We can expand  $\hat \mu_n - \mu_n$ to get another formulation of this covariance
\begin{align}
	\expect{\langle \hat \natp_n ; \hat \mu_n - \mu_n \rangle}
	&= \expect{\langle \hat \natp_n ; \frac{\sum_i T(X_i) - \mu^*}{n+n_0} \rangle}  \\
	& = \frac{n}{n+n_0} \expect{\langle \hat \natp_n ; T(X_0) - \mu^*\rangle}
	= \frac{n}{n+n_0} \Cov(\hat \natp_n ; T(X_0) ) 
\end{align}
where we used the property that all data points play a symmetric role to remove the sum over $i$.

\clearpage
